Profitability Prediction of Consulting Projects in the Construction Industry

### 2016-07-28 finalwriteup/draft.rmd
finished kerrie editing

---
title: "draft 1"
author: "Amy Cook"
date: "Tuesday, July 07, 2015"
output:
  word_document:
    reference_docx: mystyles.docx
  html_document: default
bibliography: Library.bib
---

@@

* Abstracts allow readers to quickly decide whether the research is relevant to their own interests
* get at your main argument quickly
* explain the purpose, goal methods, and results

Engaging in loss-making jobs for fixed fees is a major problem in consulting[,]
"businesses."
"This is especially" 
[particularly] 
"true" in the competitive construction industry. This thesis investigates whether machine learning techniques applied to a company's passively collected internal data could help a 
[company] avoid loss-making jobs or 
"facilitate improved commercial strategy." 
[tactfully enforce stricter contracts.] 
It was found that in a specific decision framework, a case study's profits could be improved 9% by avoiding approximately 4% of projects.
"THIS IS AWESOME!!!! WHERE CAN I GET ONE?"
Alternative decision frameworks are also 
"proposed and evaluated" 
[debated and proposed]. The internal data collected by the case study company described each of their projects in terms of employee positions, fraction of project time completed by each employee, client characteristics, project characteristics (timespan, hours, technical details) and invoicing history. Algorithmic methods such as Logistic Regression, Random Forests, Boosted Trees, Naïve Bayes, and Bayesian Networks were applied as well as blended combinations of these methods. A decision scenario 
[that] 
"which" rejected projects above a sequence of tested thresholds was run in order to find the optimal threshold for profit improvements. This process was repeated on hundreds of models built from different training subsets of the data in order to obtain a 95% statistical confidence interval of profit improvements. The blended Logistic Regression model outperformed other methods and produced a 95% confidence interval of 6.5 - 11.5% profit improvements. Further work is recommended to test user input estimation and user interface designs. The findings from this research 
"has the potential to" 
[could] assist managers in reducing losses by highlighting risky projects and 
"guiding" 
[facilitating specific changes] "improvements" to their fee structures.

* why is it important?
affects the porfitability of a company which impacts the careers, company morale, and even viability of a company.

* why should someone read your essay?

* what is your main claim?

* discuss variables and your approaches

* describe evidence to support your claim
hundred of cross validated trials using the 5 methods, as well as 5 blending methods. obtained a 95% confidence interval for each method and simple logistic regression of three models outperformed the others, closely followed by simple averaging. the 95% confidence interval across all cross validated trials was __ to __ profit improvement.

* Results: answer , was hypothesis supported
hypothesis was supprted: CAN improve bottom line of business by mobilising their passively collected data into mL algorhtms.

* implications of your work, meaning
more work to be done on testing the user interface and user ailities further. could assist managers in making better decisions about which projects they should reject or develop a more careful contract structure to improve profits.

@@

**BIM** - Building Information Modelling. A digital representation of physical characteristics of an object or structure.

@@

**CRM** - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee timesheet records. Generally accessable to each employee in a company.

@@

**FP** - False positive. A case marked 'negative' in a dataset is incorrectly predicted 'positive' by an algorithm.

@@

**FN** - False negative. A case marked 'positive' in a dataset is incorrectly predicted 'negative' by an algorithm.

**FWLS** - Feature weighted linear stacking. A linear combination of model results which are interacted with original variables. This allows certain models to be weighted heavier for certain values of the original features.

@@

**TN** - True negative. A case marked 'negative' in a dataset is correctly predicted 'negative' by an algorithm.

**TP** - True positive. The case where a case marked 'positive' in a dataset is correctly predicted 'positive' by an algorithm. Generally, 'positive' is assigned to categories for which the use case needs to be alerted.

@@

- don't mention bligh tanner!

@@

Cook, A., Wu, P., & Mengersen, K. (2015, September). Machine Learning and Visual Analytics for Consulting Business Decision Support. In Big Data Visual Analytics (BDVA), 2015 (pp. 1-2). IEEE.

@@

Signed .................
Date....................

# Chapter 1: Introduction

@@

Consulting businesses give expert advice to other professionals in exchange for a fee based on the amount of time their consultanting staff spend on the project. The deliverable 'expert advice' is generally in the form of a document or a package of documents. A document could range from a 5 page report written by a single person to a sequence of hundreds of drawings assembled by a team of 10 over many years. Despite the range of possible deliverables, before any project commences, the client and consultant must agree on a fee, or at least a fee structure. The fee structure can be as creative as the engaged parties wish, but in fact tend to fit a few common models. A couple of examples are:

1. The client agrees to pay for the consultants' time by the hour until the task is complete (referred to as hourly rate) or 
2. the consultant offers a fixed fee to complete the project in full, regardless of time spent (referred to as fixed price or fixed fee).

@@

The way a consulting manager calculates a fixed price varies from industry to industry and even company to company. Typically a consulting manager has experience in the type of project they are quoting and, after reviewing the project details, can use a combination of intuition, comparison to past projects, and rules of thumb (such as a set fraction of the entire client budget for a large project). Such qualitative methods can result in profit margins which ultimately depend on the skill (or luck) the manager has in predicting the amount of time a project demands.

This thesis focuses on a single case study consulting company in the construction industry that has persevered through many fixed price projects that resulted in losses. A source of insight into the loss-making projects was identified - twelve years of passively collected data that described each project in terms of their clients, invoice history, employee hours and technical details. Was it possible to find trends in what makes a project loss-making and can these characteristics be identified before a fixed price is offered? Can these trends be modelled into a predictive algorithm, and if so, how could the outcomes be integrated into business decisions? Finally, if the predictive outcomes were integrated into decision-making, would they deinitely improve the business' bottom line? These questions are explored in depth and answered over the course of this dissertation.

@@

# Chapter 2: Problem Description

@@

**Problem Description:**

Project managers across many industries have always struggled with forecasting project costs. A study on large-scale infrastructure projects over the past seventy years revealed that cost forecasts consistently underestimated the cost of rail projects by an average of 44.7%, the cost of bridge and tunnel projects by 33.8% and the cost of road projects by 20.4% [@Flyvbjerg2007]. A study across 1471 IT projects showed that 27% of projects ran over budget, and one in 6 of those projects were more than 200% over budget [@Flyvbjerg2011]. Another review of 6 IT project surveys between 1984 and 1994 by @Molokken2003 revealed that 60-80% of IT projects encountered effort and/or schedule overruns, where the average overrun was by 30 to 40%. These statistics describe difficult conditions that can result in job losses, business bankruptcies, and government budget blow outs. A survey from @Moores1992 asked software managers whether they saw effort estimation as a problem, and 91% responded 'yes'.

@@

**A Wicked Problem**

The challenge for consultants to reduce budget blowouts, in terms of hours spent, is not easily solved. Complex projects always differ from one another - even similar projects may end up demanding significantly different amounts of time. Although this is known, a fixed price is often demanded by clients before a project is awarded. If an hourly rate fee structure is proposed, clients in many industries can simply turn to another consultant willing to take the financial risks involved with fixed prices **(...ref...)**. Two strategies to guard against fixed price budget blowouts are detailed below along with their limitations.

A compromise between the hourly rate and fixed price structure is to track employee hours, and stop work once the fixed consulting fee has been exhausted. Then, negotiate further fees, or variations, with the client. This can either be agreed upon with the client beforehand or the consultant's internal strategy. If this strategy is proposed to the client up front, it is likely the client would again turn to another consultant who is willing to offer a fixed price. If this approach is the consultant's internal strategy, their financial risk may satisfactorily be reduced. However, the behaviour may cause friction between the client and the consultant, therefore reducing the potential for repeat business. In the event of hours inflating beyond expectation, consultants may choose to wear losses or endure marginable profits to preserve their client relationship. They may reason it is in their best interest to maintain a reliable, trustworthy reputation in their industry above creating financial friction.

@@

Clearly, there is not a straightforward contractual solution to reducing the financial risk taken on by consultant's who offer fixed prices in complex projects. In many industries, the reality of fixed price fee structures are unavoidable and businesses conform to survive.

**Current Fixed Price Estimation Methods:**

It has been established that estimating fixed price fees in consulting is a difficult task with unfortunate consequences if the price is estimated too conservatively. Thie project aims to mobilise a consultant's historic data describing profitable and loss-making jobs in order to improve their fixed price setting practices. But what is the current method for estimating fixed prices or project costs? Is past data currently used in other cost estimation practises and to what degree of success?

@@

**Cost Estimation and Human Nature**

Although the surveys indicated industry did not analyse past project cost data, the idea of using this data (in the same way) has been studied before. Research exists that tested people's estimation accuracy with and without comparitive historic data with interesting results. This is summarised below along with the resulting data-driven models.

@Lovallo2003 delved into the psychology behind why executives so often severely underestimate costs of larger projects such as manufacturing plant construction, mergers and acquisitions, large infrastructure and software development. Their theory stemmed from Kahneman's work on decision-making that won him the Nobel Prize for economics in 2002. His research argued that a person's natural optimistic view of their own skills leads to consistent underestimation of the time and risks involved in a project. A manager optimistically sees challenges in a project as something that can be overcome by the team's high skill level, and downplays or ignores the risk of problems that are out of the team's control. This is why it does not matter if the project is broken down to the highest level of detail for cost prediction, all complex projects are at risk of encountering a multitude of problems that the manager could never foresee. Each problem has a low chance of occurring, but in combination the risk is much greater [@Lovallo2003].

@Lovallo2003 call the practice of analysing a project based on project details without considering unforseen risks the 'inside view'. A survey by @Moores1992 aligned with this theory and found the most popular response for the supposed reason for cost overruns was over-optimistic estimates (51%). This contrasts with an 'outside view' where a subject is analysed by comparison to other subjects and disregarding specific subject details. For example, research has shown that if people make predictions about their skills after being exposed to an 'outside view' (a summary of other people's skills), their predictions are significantly more accurate. According to @Lovallo2003, this 'outside view' can be applied to complex projects, and stipulates that the details of the current project should be ignored in favour of analysing outcomes of several similar projects. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommended was to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model, which is then used to adjust the forecasted cost made by detailed analysis (the 'inside view') [@Flyvbjerg2011]. As demonstrated in the surveys, @Flyvbjerg2011 found that managers rarely analysed data from similar projects even though it could significantly improve cost predictions. This discovery led the team to formally introduce 'outside view' methods into some institutions.

The implementation of reference class forecasting began in project management for the first time in 2004. It was endorsed by the American Planning Association in 2005, and is now used in some governments and private companies in Europe, South Africa, and Australia. An example of the type of output from a reference class statistical model is a plot showing the relationship between the acceptable chance of cost overrun and the required uplift to the original forecasted cost [@Flyvbjerg2011]. An example reading from the chart is as follows: "If the chance of a cost overrun of 20% is acceptable, the estimated cost should be increased by 55% from original calculations.". This builds past experiences of cost overruns into the initial cost estimation.

@@

The visualisation presents a powerful communication tool for influencing decision makers and improving forecast cost accuracy, however, access to credible data for a sufficient number of projects can be a challenge [@Flyvbjerg2011]. Overall, the optimism bias theory credibly helps to explain the persistant problem of project cost underestimation.

@@

Although @Flyvbjerg2011's reference class forecasting theory and the psychology behind it was first published in 2003, predicting project costs using previous data has been researched for many preceding years and continues to this day. In this section, a brief overview of other reference class forecasting models is presented (they are reviewed in greater detail in the literature review) followed by an analysis of why industry has not yet adopted them.

Most cost estimation or reference forecasting analytical models focused on software development and construction projects, along with some work on film projects. These analyses generally used information from 15 to 20 large projects and the dataset summarised projects from around the world with different clients and teams. Applied mathematical methods varied from statistical Regressions to deep-learning Neural Networks [@Love2005][@Kim2004]. Most studies reported that their predictive models performed well and that Neural Networks were generally more accurate than Linear Regression.

@@

In the software industry it is also possible the problem of fixed price contract blowouts has been solved by an alternative contract structure which has made cost estimation models unnecessary. In contrast to construction, IT projects are a relatively new practice and the industry has had success in trialing alternative contractural arangements. Through experimentation, the Agile movement was developed. In summary, the method treats both cost and time as fixed quantities for a project, and any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted and revised based on frequent progress review points that assess unexpected problems. The flexible upfront arrangement with clients has often been found to forge better long term relationships [Cockburn2001]. This agile methodology suits the nature of software development work, where curtailing the scope of work is relatively straightforward. Also, since the industry is more modern, the new methodology was able to sweep across the field and could be proven to deliver better budget and product outcomes to clients [@Rasmusson2010].

Unfortunately, in other industries such as the construction industry, infrastructure and building projects are steeped in a traditional process, where contractual norms have developed over a much longer time period [@Badenfelt2011]. The industry is so large, broad and well established that it has inertia in the way projects are estimated, and contractually bound. Furthermore, the industry would struggle to take on an agile approach as the scope cannot be easily reduced as the project unfolds - the project must result in construction and building certification which implies a high level of legal responsibility and due diligence.

@@

The case study company is an engineering consulting company in the construction industry. In their profession, labour is the chief cost and the long traditional history of this industry mandates fixed price projects as the norm. In recent years, the size of losses from unprofitable projects equalled 25-30% of their profits from profitable projects. This demonstrates the magnitude of their cost estimation problem and the substantial room for improvement that a data-driven model may facilitate.

Cost estimation is currently performed in the case study company by first carefully reviewing preliminary drawings where time costs for each task are estimated and summed by an experienced manager (the 'inside view'). This is often cross-checked with a value based off a percentage of the estimated final cost of construction (generally the client budget). No formal mathematical method of comparison to similar projects is performed. This is the case for a number of possible reasons:

* lack of time to interrogate past project data
* clunky data availability offered by their database software
* lack of awareness of similar projects performed by other managers

A Customer Relationship Management software package (CRM) is currently employed to collect and store project data. A CRM is a popular type of software used by businesses to record client and project details as well as employee timesheet records. In the case study business, the CRM is available to all employees over the company intranet, and each employee completes daily timesheets alotting their hours to certain projects. Additionally, technical information is recorded against each project as well and the CRM data stood as an untapped source of information within the case study organisation. It stores a rich variety of data including:

* employee timesheet hours with dates
* other project costs (taxis, printing)
* client information/characteristics
* client identification code
* invoiced amounts for each project and dates
* employee costs
* employee charge out rates
* project description

The CRM software readily provides simple output statistics, however, analytical capabilities were limited to:

* simple scatter plots and bar charts of the raw data
* summaries such as overall hours spent vs. the invoiced amount for each project

The company had a wealth of data but limited means to extract insight. This study exploits the CRM data by performing more sophisticated statistical analysis with the intention of building a predictive model to improve cost estimation. Thousands of past projects are available for analysis, as opposed to previous studies that used on average 15-20 cases and at most 300 [@Finnie1997][@Pai2013][@Shin2015]. This improves the potential for accurate predictions. If successful, the model also has higher potential for managerial uptake as the algorithm will be directly built and trained on internal company data. Managers could interrogate how the model outputs were calculated and relate to the actual projects used in the model or speak to a colleague that was involved. This may give the presented 'outside view' a stronger meaning and more potential to influence the final fee proposal.

@@

The most obvious limitation is that the work focuses on a single company in a specific industry, which limits the application of conclusions to a broader range of companies. This can be overcome by first determining the optimum method and variables for this case and then testing them on future case study companies.

@@

Finally, all CRM data is entered by employees manually. Because the data is input by people, it is susceptible to some error. Many errors were detectable during data cleaning, however it is possible that a low number of undetectable human errors exist in the data and marginally influence predictions.

@@

In comparison to previous studies, this project advances the body of work on cost estimation in three ways: the dataset is larger than other studies and sourced from a single company's CRM database, ensemble tree methods and model blending was tested, and the prediction results were integrated into a decision framework for the business. These points are outlined in more detail below.

In past research, cost estimation datasets generally consisted of dozens of projects from a range of a companys and even countries. This thesis case study only had access to the company's internal data but this contained over 2,000 past jobs - far more than the usual cost estimation study and with many more variables. Furthermore, the problem of cost estimation is applied to a consulting company in the construction industry which differs from other studies in the construction industry that estimate actual construction costs. Consulting in the construction industry can be likened more to software development effort estimation although the nature of the work differs considerably.

In addition to addressing cost estimation with a different kind of dataset, ensemble tree methods are applied which have been minimally tested on this problem. Ensemble tree methods combine hundreds or thousands of decision trees to make predictions. A single decision tree is a non-parametric machine-learned model built by progressively determining the best binary split to partition the data according to the response variable. For further explanation of the method behind decision trees and ensemble three excellent sources are @Hastie2009, @Elith2008, and @Breiman2005. To date, Linear Regression, Neural Networks, Case Based Reasoning, and Support Vector Machines (SVM's) have been used to predict cost/effort even though ensemble tree methods can perform as well as Neural Networks in some cases and generally outperform Linear Regression (**ref**). Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots that provide insight into the model's calculations. This contrasts Neural Networks and SVM's which are 'black-box' predictors. Providing insight to a user is a major benefit as past research has demonstrated issues with user trust and uptake of models, even if the models predicted well. It is anticipated that using internal data will also improve trust and create tangible meaning for the user. Blending multiple machine learning and statistical models has also not yet been applied to the cost estimation problem and may improve predictions.

Finally, this study propels the predictive model one step further than other studies by analysing bottom-line profit improvements that could result from the research. A few possibile ways to integrate model results into business decision-making scenarios are presented, and where possible, final profit increases are evaluated. This kind of thought experiment is designed to present a clear case for industry when evaluating if and how to adopt cost estimation models.

@@

Chapter 4 describes the method followed throughout the research process. It includes first how the data was obtained, the lengthy cleaning process, followed by variable importance analysis, variable selection, trials of selected predictive algorithms and problem constructs. The cost estimation problem was attempted first as a regression problem predicting return per dollar, followed by a simpler binary classification problem predicting profit or loss. Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision making scenarios. Further applications were also considered.

>>>>

---
title: "Profitability Prediction of Consulting Projects in the Construction Industry"
author: "Amy Cook"
date: "Tuesday, July 07, 2015"
output:
  word_document:
    reference_docx: mystyles.docx
  html_document: default
bibliography: Library.bib
csl: apa.csl
---

@@

Engaging in loss making jobs for fixed fees is a major problem in consulting, particularly in the competitive construction industry. This thesis investigates whether machine learning techniques applied to a company's passively collected internal data could help avoid loss making jobs tactfully choose when to enforce stricter contracts. It was found that in a specific decision framework, a case study's profits could be improved 9% by avoiding approximately 4% of projects. Alternative decision frameworks are also proposed and evaluated. The internal data collected by the case study company described each of their projects in terms of employee positions, fraction of project time completed by each employee, client characteristics, project characteristics (time span, hours, technical details) and invoicing history. Algorithmic methods such as Logistic Regression, Random Forests, Boosted Trees, Naive Bayes, and Bayesian Networks were applied as well as blended combinations of these methods. A decision scenario which" rejected projects above a sequence of tested thresholds was run in order to find the optimal threshold for profit improvements. This process was repeated on hundreds of models built from different training subsets of the data in order to obtain a 95% statistical confidence interval of profit improvements. The blended Logistic Regression model outperformed other methods and produced a 95% confidence interval of 6.5 - 11.5% profit improvements. Further work is recommended to test user input estimation and user interface designs. The findings from this research have the potential to assist managers in reducing losses by highlighting risky projects and guiding project-based changes to fee structures.

@@

**BIM** - Building Information modeling. A digital representation of physical characteristics of an object or structure.

@@

**CRM** - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee time sheet records. Generally accessible to each employee in a company.

@@

**FP** - False positive. A case marked 'negative' in a data set is incorrectly predicted 'positive' by an algorithm.

@@

**FN** - False negative. A case marked 'positive' in a data set is incorrectly predicted 'negative' by an algorithm.

**FWLS** - Feature weighted linear stacking. A linear combination of model results that is interacted with original variables. This allows certain models to be weighted heavier for certain values of the original features.

@@

**TN** - True negative. A case marked 'negative' in a data set is correctly predicted 'negative' by an algorithm.

**TP** - True positive. The case where a case marked 'positive' in a data set is correctly predicted 'positive' by an algorithm. Generally, 'positive' is assigned to categories for which the use case needs to be alerted.

@@

I would like to express special appreciation to my Principal Supervisor Professor Kerrie Mengersen. Thanks for extending a hand and giving me this opportunity as well as deftly guiding my work and welcoming me into your excellent research environment. Next, I would like to thank my Associate Supervisor Dr. Paul Wu for kindly and cheerfully working with me on the finer details of my research, giving up many hours of your time. Lastly, I would not have learned half as much without my colleagues in the BRAG group and ACEMS team.

@@

Cook, A., Wu, P., & Mengersen, K. (2015, September). Machine Learning and Visual Analytics for Consulting Business Decision Support. In *Big Data Visual Analytics (BDVA), 2015* (pp. 1-2). IEEE.

@@

Signed ................................
Date....................

# Introduction

@@

Consulting businesses give expert advice to other professionals in exchange for a fee based on the amount of time their consulting staff spends on the project. The deliverable 'expert advice' is generally in the form of a document or a package of documents. A document could range from a 5-page report written by a single person to a sequence of hundreds of drawings assembled by a team of 10 over many years. Despite the range of possible deliverables, before any project commences, the client and consultant must agree on a fee, or at least a fee structure. The fee structure can be as creative as the engaged parties wish, but in fact tend to fit a few common models. A couple of examples are:

1. The client agrees to pay for the consultants' time by the hour until the task is complete (referred to as hourly rate) or 
2. The consultant offers a fixed fee to complete the project in full, regardless of time spent (referred to as fixed price or fixed fee).

@@

The way a consulting manager calculates a fixed price varies from industry to industry and even company to company. Typically a consulting manager has experience in the type of project they are quoting and, after reviewing the project details, can use a combination of intuition, comparison to past projects, and rules of thumb (such as a set fraction of the entire client budget for a large project). Such qualitative methods can result in profit margins, which ultimately depend on the skill (or luck) the manager has in predicting the amount of time a project demands.

This thesis focuses on a single case study consulting company in the construction industry that has persevered through many fixed price projects that resulted in losses. A source of insight into the loss making projects was identified - twelve years of passively collected data that described each project in terms of their clients, invoice history, employee hours and technical details. Was it possible to find trends in what makes a project loss making and can these characteristics be identified before a fixed price is offered? Can these trends be modeled into a predictive algorithm, and if so, how could the outcomes be integrated into business decisions? Finally, if the predictive outcomes were integrated into decision-making, would they definitely improve the business' bottom line? These questions are explored in depth and answered over the course of this dissertation.

@@

# Problem Description

@@

### Problem Description

Project managers across many industries have always struggled with forecasting project costs. A study on large-scale infrastructure projects over the past seventy years revealed that cost forecasts consistently underestimated the cost of rail projects by an average of 44.7%, the cost of bridge and tunnel projects by 33.8% and the cost of road projects by 20.4% [@Flyvbjerg2007]. A study across 1471 IT projects showed that 27% of projects ran over budget, and one in 6 of those projects were more than 200% over budget [@Flyvbjerg2011]. Another review of 6 IT project surveys between 1984 and 1994 by @Molokken2003 revealed that 60-80% of IT projects encountered effort and/or schedule overruns, where the average overrun was by 30 to 40%. These statistics describe difficult conditions that can result in job losses, business bankruptcies, and government budget blowouts. A survey from @Moores1992 asked software managers whether they saw effort estimation as a problem, and 91% responded 'yes'.

@@

### A Wicked Problem

The challenge for consultants to reduce budget blowouts, in terms of hours spent, is not easily solved. Complex projects always differ from one another - even similar projects may end up demanding significantly different amounts of time. Although this is known, clients often request a fixed price before a project is awarded. If an hourly rate fee structure is proposed, clients in many industries can simply turn to another consultant willing to take the financial risks involved with fixed prices [@Pai2013]. Two strategies to guard against fixed price budget blowouts are detailed below along with their limitations.

A compromise between the hourly rate and fixed price structure is to track employee hours, and stop work once the fixed consulting fee has been exhausted. Then, negotiate further fees, or variations, with the client. This can either be agreed upon with the client beforehand or the consultant's internal strategy. If this strategy is proposed to the client up front, it is likely the client would again turn to another consultant who is willing to offer a fixed price. If this approach is the consultant's internal strategy, their financial risk may satisfactorily be reduced. However, the behaviour may cause friction between the client and the consultant, therefore reducing the potential for repeat business. In the event of hours inflating beyond expectation, consultants may choose to wear losses or endure marginal profits to preserve their client relationship. They may reason it is in their best interest to maintain a reliable, trustworthy reputation in their industry above creating financial friction.

@@

Clearly, there is not a straightforward contractual solution to reducing the financial risk taken on by consultant's who offer fixed prices in complex projects. In many industries, the reality of fixed price fee structures is unavoidable and businesses conform to survive.

### Current Fixed Price Estimation Methods

It has been established that estimating fixed price fees in consulting is a difficult task with unfortunate consequences if the price is estimated too conservatively. This project aims to mobilise a consultant's historic data describing profitable and loss making jobs in order to improve their fixed price setting practices. But what is the current method for estimating fixed prices or project costs? Is past data currently used in other cost estimation practices and to what degree of success?

@@

### Cost Estimation and Human Nature

Although the surveys indicated industry did not analyse past project cost data, the idea of using this data (in the same way) has been studied before. Research exists that tested people's estimation accuracy with and without comparative historic data with interesting results. This is summarised below along with the resulting data-driven models.

@Lovallo2003 delved into the psychology behind why executives so often severely underestimate costs of larger projects such as manufacturing plant construction, mergers and acquisitions, large infrastructure and software development. Their theory stemmed from Kahneman's work on decision-making that won him the Nobel Prize for economics in 2002. His research argued that a person's natural optimistic view of their own skills leads to consistent underestimation of the time and risks involved in a project. A manager optimistically sees challenges in a project as something that can be overcome by the team's high skill level, and downplays or ignores the risk of problems that are out of the team's control. This is why it does not matter if the project is broken down to the highest level of detail for cost prediction; all complex projects are at risk of encountering a multitude of problems that the manager could never foresee. Each problem has a low chance of occurring, but in combination the risk is much greater [@Lovallo2003].

@Lovallo2003 called the practice of analysing a project based on project details without considering unforeseen risks the 'inside view'. A survey by @Moores1992 aligned with this theory and found the most popular response for the supposed reason for cost overruns was over-optimistic estimates (51%). This contrasts with an 'outside view' where a subject is analysed by comparison to other subjects and disregarding specific subject details. For example, research has shown that if people make predictions about their skills after being exposed to an 'outside view' (a summary of other people's skills), their predictions are significantly more accurate. According to @Lovallo2003, this 'outside view' can be applied to complex projects, and stipulates that the details of the current project should be ignored in favour of analysing outcomes of several similar projects. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommended was to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model, which is then used to adjust the forecast cost made by detailed analysis (the 'inside view') [@Flyvbjerg2011]. As demonstrated in the surveys, @Flyvbjerg2011 found that managers rarely analysed data from similar projects even though it could significantly improve cost predictions. This discovery led the team to formally introduce 'outside view' methods into some institutions.

The implementation of reference class forecasting began in project management for the first time in 2004. It was endorsed by the American Planning Association in 2005, and is now used in some governments and private companies in Europe, South Africa, and Australia. An example of the type of output from a reference class statistical model is a plot showing the relationship between the acceptable chance of cost overrun and the required uplift to the original forecast cost [@Flyvbjerg2011]. An example reading from the chart is as follows: "If the chance of a cost overrun of 20% is acceptable, the estimated cost should be increased by 55% from original calculations.". This builds past experiences of cost overruns into the initial cost estimation.

@@

The visualisation presents a powerful communication tool for influencing decision makers and improving forecast cost accuracy, however, access to credible data for a sufficient number of projects can be a challenge [@Flyvbjerg2011]. Overall, the optimism bias theory credibly helps to explain the persistent problem of project cost underestimation.

@@

Although Flyvbjerg's (2011) reference class forecasting theory and the psychology behind it was first published in 2003, predicting project costs using previous data has been researched for many preceding years and continues to this day. In this section, a brief overview of other reference class forecasting models is presented (they are reviewed in greater detail in the literature review) followed by an analysis of why industry has not yet adopted them.

Most cost estimation or reference forecasting analytic models focused on software development and construction projects, along with some work on film projects. These analyses generally used information from 15 to 20 large projects and the data set summarised projects from around the world with different clients and teams. Applied mathematical methods varied from statistical Regressions to deep-learning Neural Networks [@Love2005; @Kim2004]. Most studies reported that their predictive models performed well and that Neural Networks were generally more accurate than Linear Regression.

@@

In the software industry it is also possible the problem of fixed price contract blowouts has been solved by an alternative contract structure, which has made cost estimation models unnecessary. In contrast to construction, IT projects are a relatively new practice and the industry has had success in trialing alternative contractual arrangements. Through experimentation, the Agile movement was developed. In summary, the method treats both cost and time as fixed quantities for a project, and any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted and revised based on frequent progress review points that assess unexpected problems. The flexible upfront arrangement with clients has often been found to forge better long-term relationships [Cockburn2001]. This agile methodology suits the nature of software development work, where curtailing the scope of work is relatively straightforward. Also, since the industry is more modern, the new methodology was able to sweep across the field and could be proven to deliver better budget and product outcomes to clients [@Rasmusson2010].

Unfortunately, in other industries such as the construction industry, infrastructure and building projects are steeped in a traditional process, where contractual norms have developed over a much longer time period [@Badenfelt2011]. The industry is so large, broad and well established that it has inertia in the way projects are estimated, and contractually bound. Furthermore, the industry would struggle to take on an agile approach as the scope cannot be easily reduced as the project unfolds - the project must result in construction and building certification which implies a high-level of legal responsibility and due diligence.

@@

The case study company is an engineering consulting company in the construction industry. In their profession, labour is the chief cost and the long traditional history of this industry mandates fixed price projects as the norm. In recent years, the size of losses from unprofitable projects equaled 25-30% of their profits from profitable projects. This demonstrates the magnitude of their cost estimation problem and the substantial room for improvement that a data-driven model may facilitate.

Cost estimation is currently performed in the case study company by first carefully reviewing preliminary drawings where time costs for each task are estimated and summed by an experienced manager (the 'inside view'). This is often crosschecked with a value based off a percentage of the estimated final cost of construction (generally the client budget). No formal mathematical method of comparison to similar projects is performed. This is the case for a number of possible reasons:

* Lack of time to interrogate past project data
* Clunky data availability offered by their database software
* Lack of awareness of similar projects performed by other managers

A Customer Relationship Management software package (CRM) is currently employed to collect and store project data. A CRM is a popular type of software used by businesses to record client and project details as well as employee time sheet records. In the case study business, the CRM is available to all employees over the company intranet, and each employee completes daily time sheets allotting their hours to certain projects. Additionally, technical information is recorded against each project as well and the CRM data stood as an untapped source of information within the case study organisation. It stores a rich variety of data including:

* Employee time sheet hours with dates
* Other project costs (taxis, printing)
* Client information/characteristics
* Client identification code
* Invoiced amounts for each project and dates
* Employee costs
* Employee charge out rates
* Project description

The CRM software readily provides simple output statistics, however, analytic capabilities were limited to:

* Simple scatter plots and bar charts of the raw data
* Summaries such as overall hours spent vs. the invoiced amount for each project

The company had a wealth of data but limited means to extract insight. This study exploits the CRM data by performing more sophisticated statistical analysis with the intention of building a predictive model to improve cost estimation. Thousands of past projects are available for analysis, as opposed to previous studies that used on average 15-20 cases and at most 300 [@Finnie1997; @Pai2013; @Shin2015]. This improves the potential for accurate predictions. If successful, the model also has higher potential for managerial uptake as the algorithm will be directly built and trained on internal company data. Managers could interrogate how the model outputs were calculated and relate to the actual projects used in the model or speak to a colleague that was involved. This may give the presented 'outside view' a stronger meaning and more potential to influence the final fee proposal.

@@

The most obvious limitation is that the work focuses on a single company in a specific industry, which limits the application of conclusions to a broader range of companies. First determining the optimum method and variables for this case and then testing them on future case study companies can overcome this.

@@

Finally, employees enter all CRM data manually. Because the data is input by people, it is susceptible to some error. Many errors were detectable during data cleaning, however it is possible that a low number of undetectable human errors exist in the data and marginally influence predictions.

@@

In comparison to previous studies, this project advances the body of work on cost estimation in three ways: the data set is larger than other studies and sourced from a single company's CRM database, ensemble tree methods and model blending was tested, and the prediction results were integrated into a decision framework for the business. These points are outlined in more detail below.

In past research, cost estimation data sets generally consisted of dozens of projects from a range of a companies and even countries. This thesis case study only had access to the company's internal data but this contained over 2,000 past jobs - far more than the usual cost estimation study and with many more variables. Furthermore, the problem of cost estimation is applied to a consulting company in the construction industry, which differs from other studies in the construction industry that estimate actual construction costs. Consulting in the construction industry can be likened more to software development effort estimation although the nature of the work differs considerably.

In addition to addressing cost estimation with a different kind of data set, ensemble tree methods are applied which have been minimally tested on this problem. Ensemble tree methods combine hundreds or thousands of decision trees to make predictions. A single decision tree is a non-parametric machine-learned model built by progressively determining the best binary split to partition the data according to the response variable. For further explanation of the method behind decision trees and ensemble trees, three excellent sources are @Hastie2009, @Elith2008, and @Breiman2005. To date, Linear Regression, Neural Networks, Case Based Reasoning, and Support Vector Machines (SVM's) have been used to predict cost/effort even though ensemble tree methods can perform as well as Neural Networks and generally outperform Linear Regression [@Caruana2006]. Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots that provide insight into the model's calculations. This contrasts Neural Networks and SVM's, which are 'black-box' predictors. Providing insight to a user is a major benefit as past research has demonstrated issues with user trust and uptake of models, even if the models predicted well. It is anticipated that using internal data will also improve trust and create tangible meaning for the user. Blending multiple machine learning and statistical models has also not yet been applied to the cost estimation problem and may improve predictions.

Finally, this study propels the predictive model one step further than other studies by analysing bottom-line profit improvements that could result from the research. A few possible ways to integrate model results into business decision-making scenarios are presented, and where possible, final profit increases are evaluated. This kind of thought experiment is designed to present a clear case for industry when evaluating if and how to adopt cost estimation models.

@@

Chapter 4 describes the method followed throughout the research process. It includes first how the data was obtained, the lengthy cleaning process, followed by variable importance analysis, variable selection, trials of selected predictive algorithms and problem constructs. The cost estimation problem was attempted first as a regression problem predicting return per dollar, followed by a simpler binary classification problem predicting profit or loss. Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision-making scenarios. Further applications were also considered.

The research outlined in the method was divided into four stages, each of which was assigned a chapter containing the results and discussion of that portion. Chapter 5 describes the variable selection process. Chapter 6 and 7 cover the results of the two ways of framing the cost estimation problem: using regression models and binary classification models. Chapter 8 shows the outcomes of model blending, and then Chapter 9 extends the predictive analyses into an overall profit analysis, as well as describing research limitations and future work.
